---
title: "Deconvolve Study1 microarray"
author: "Penghui Huang"
date: "2024-10-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## CTB

```{r}
ctb<-read.delim("/storage/windows-backup/D drive/Penghui/collab/Study 1 data/ctbmbparray.txt",header=TRUE)
ctb <- ctb[ctb$Gene != "NOGENE",]
keep.gene <- names(table(ctb$Gene)[which(table(ctb$Gene) == 12)])
ctb <- ctb[ctb$Gene %in% keep.gene,]
dose<-factor(ctb$dose2, level=c(1,2,3,4), ordered=TRUE)
genenumvec <- unique(ctb$genenum)
ngenes <- length(genenumvec)
genename<-as.character(ctb$Gene)
genename <- unique(genename)

xlabel <- unique(ctb$X_LABEL_)
ctb_bulk <- list()
for(i in 1:length(xlabel)){
  tmp <- ctb[ctb$X_LABEL_ == xlabel[i],]
  rownames(tmp) <- tmp$Gene
  tmp <- tmp[genename,]
  ctb_bulk <- c(ctb_bulk, list(as.vector(tmp$log2value)))
}
ctb_bulk <- rlist::list.cbind(ctb_bulk)
rownames(ctb_bulk) <- genename
colnames(ctb_bulk) <- xlabel
```

## Reference Prep
```{r}
load("/storage/windows-backup/D drive/Penghui/collab/ref_deconv/Broad.RData")
load("/storage/windows-backup/D drive/Penghui/collab/ref_deconv/Suryawanshi_2018.RData")
load("/storage/windows-backup/D drive/Penghui/collab/ref_deconv/Campbell_2022.RData")
load("/storage/windows-backup/D drive/Penghui/collab/ref_deconv/Liu_2018.RData")
Derisoud <- readxlsb::read_xlsb("/storage/windows-backup/D drive/Penghui/collab/Cell Deconvolution/Derisoud_2024/Supplementary Table S8.xlsb", sheet = "Average exp term")
# There are some genes that occur twice or three times in the reference
# These rows are identical
# Only keep one occurence
Derisoud <- unique.data.frame(Derisoud)
rownames(Derisoud) <- Derisoud$Shee
Derisoud <- as.matrix(Derisoud[,-1])

colnames(Broad)
colnames(Campbell_2022_cts)
colnames(Liu_2018_cts)
colnames(Suryawanshi_2018_cts)
colnames(Derisoud)
```

```{r}
# In total, 7 cell types

Cytotrophoblasts <- list("Broad"=Broad[,"vCTB"],
                         "Campbell"=rowMeans(Campbell_2022_cts[,c("Cytotrophoblasts","Proliferative Cytotrophoblasts")]),
                         "Liu"=Liu_2018_cts[,"CTB"],
                         "Suryawanshi"=Suryawanshi_2018_cts[,"VCT"])
Extravillous_Trophoblasts <- list("Broad"=Broad[,"EVT"],
                         "Liu"=Liu_2018_cts[,"EVT"],
                         "Suryawanshi"=Suryawanshi_2018_cts[,"EVT"])
Syncytiotrophoblasts <- list("Broad"=Broad[,"STB"],
                           "Campbell"=Campbell_2022_cts[,"Syncytiotrophoblast"],
                           "Liu"=Liu_2018_cts[,"STB"],
                           "Suryawanshi"=Suryawanshi_2018_cts[,"SCT"])
Hofbauer <- list("Broad"=Broad[,"Hofbauer cells"],
                 "Campbell"=Campbell_2022_cts[,"Hofbauer Cells"],
                 "Suryawanshi"=Suryawanshi_2018_cts[,"HC"])
Stromal <- list("Broad"=Broad[,"Fibroblast"],
                "Campbell"=Campbell_2022_cts[,"Fibroblasts"],
                "Liu"=Liu_2018_cts[,"STR"],
                "Suryawanshi"=Suryawanshi_2018_cts[,"FB"])
Perivascular <- list("Suryawanshi"=Suryawanshi_2018_cts[,"VEC"],
                     "Suryawanshi_rep"=Suryawanshi_2018_cts[,"VEC"])
Endothelial <- list("Broad"=Broad[,"Endothelial_cells"],
                    "Campbell"=Campbell_2022_cts[,"Endothelial Cells"],
                    "Suryawanshi"=Suryawanshi_2018_cts[,"VEC"])
```


```{r}
genes.shared <- Reduce(intersect, lapply(Cytotrophoblasts, names))
Cytotrophoblasts <- lapply(Cytotrophoblasts, function(x){x[genes.shared]})
Extravillous_Trophoblasts <- lapply(Extravillous_Trophoblasts, function(x){x[genes.shared]})
Syncytiotrophoblasts <- lapply(Syncytiotrophoblasts, function(x){x[genes.shared]})
Hofbauer <- lapply(Hofbauer, function(x){x[genes.shared]})
Stromal <- lapply(Stromal, function(x){x[genes.shared]})
Perivascular <- lapply(Perivascular, function(x){x[genes.shared]})
Endothelial <- lapply(Endothelial, function(x){x[genes.shared]})

Cytotrophoblasts <- rlist::list.cbind(Cytotrophoblasts)
Extravillous_Trophoblasts <- rlist::list.cbind(Extravillous_Trophoblasts)
Syncytiotrophoblasts <- rlist::list.cbind(Syncytiotrophoblasts)
Hofbauer <- rlist::list.cbind(Hofbauer)
Stromal <- rlist::list.cbind(Stromal)
Perivascular <- rlist::list.cbind(Perivascular)
Endothelial <- rlist::list.cbind(Endothelial)
```


```{r}
ref.list <- list("Cytotrophoblasts"=Cytotrophoblasts,
                 "Extravillous_Trophoblasts"=Extravillous_Trophoblasts,
                 "Syncytiotrophoblasts"=Syncytiotrophoblasts,
                 "Hofbauer"=Hofbauer,
                 "Stromal"=Stromal,
                 "Perivascular"=Perivascular,
                 "Endothelial"=Endothelial)
```


```{r}
length(intersect(rownames(ctb_bulk), rownames(ref.list[[1]])))
```



```{r}
ctb_bulk_million <- 2^ctb_bulk
for(i in 1:ncol(ctb_bulk_million)){
  ctb_bulk_million[,i] <- (ctb_bulk_million[,i]/sum(ctb_bulk_million[,i]))*1000000
}
ctb_bulk_million <- round(ctb_bulk_million)
```


```{r}
library(BLEND)
ctb_res <- BLEND(bulk = ctb_bulk_million,
                 phi = ref.list,
                 alpha = 1.0001,
                 beta=1.0001,
                 ncore = 10,
                 method = "EMMAP")
ctb_frac <- rlist::list.rbind(lapply(ctb_res, function(x){x[[1]]}))
ctb_frac
```

```{r}
Derisoud <- readxlsb::read_xlsb("/storage/windows-backup/D drive/Penghui/collab/Cell Deconvolution/Derisoud_2024/Supplementary Table S8.xlsb", sheet = "Average exp term")
Derisoud <- unique.data.frame(Derisoud)
rownames(Derisoud) <- Derisoud[,1]
Derisoud <- Derisoud[,-1]
```

```{r}
library(EpiDISH)
inter.gene <- intersect(rownames(Derisoud), rownames(ctb_bulk))
Derisoud.epidish <- Derisoud[inter.gene,]
ctb_bulk.epidish <- ctb_bulk[inter.gene,]
Derisoud.epidish <- apply(Derisoud.epidish, 2, function(x){(x/sum(x))*10000})
ctb_bulk.epidish <- apply(ctb_bulk.epidish, 2, function(x){(x/sum(x))*10000})
res_epidish <- epidish(beta.m = (as.matrix(ctb_bulk.epidish)),
                       ref.m = as.matrix(Derisoud.epidish),
                       method = "RPC")
summary(res_epidish$estF)
```


```{r}
round(apply(res_epidish$estF, 2, mean),3)
round(apply(res_epidish$estF, 2, sd),3)
```



First, the distribution of CTB microarray data look too normal. It is not skewed as
sequencing data. It is weird. 

I tried both BLEND and epidish to deconvolve microarray data. Neither can provided
pure cell type. And the major cell type is not even CTB. Microarray data like this 
may not be able to be deconvolved using regular bulk data deconvolution methods.





## TBPC



```{r}
tbpc <- read.table("/storage/windows-backup/D drive/Penghui/collab/Study 1 data/tbpcmbparray.txt", header=T)
meta <- unique.data.frame(tbpc[,4:6])
head(meta)
```


```{r}
sample.id <- unique(meta$X_LABEL_)
genes <- table(tbpc$Gene)
genes <- genes[genes == 15] 
genes <- names(genes)
bulk.list <- list()
for(i in 1:length(sample.id)){
  tmp <- tbpc[tbpc$X_LABEL_ == sample.id[i],c("Gene","log2value")]
  tmp <- tmp[tmp$Gene %in% genes,]
  rownames(tmp) <- tmp$Gene
  tmp <- tmp[genes,]
  tmp <- tmp$log2value
  bulk.list <- c(bulk.list,list(tmp))
}
tbpc.bulk <- rlist::list.cbind(bulk.list)
rownames(tbpc.bulk) <- genes
colnames(tbpc.bulk) <- sample.id
# put it back to linear scale
tbpc.bulk <- 2^tbpc.bulk
tbpc.bulk.round <- round(tbpc.bulk)
```

```{r}
# Deconvolution with BLEND
tbpc_res <- BLEND(bulk = tbpc.bulk.round,
                 phi = ref.list,
                 alpha = 1.0001,
                 beta=1.0001,
                 ncore = 10,
                 method = "EMMAP")
tbpc_frac <- rlist::list.rbind(lapply(tbpc_res, function(x){x[[1]]}))
summary(tbpc_frac)
```

```{r}
inter.gene <- intersect(rownames(Derisoud), rownames(tbpc.bulk))
Derisoud.epidish <- Derisoud[inter.gene,]
tbpc_bulk.epidish <- tbpc.bulk[inter.gene,]
Derisoud.epidish <- apply(Derisoud.epidish, 2, function(x){(x/sum(x))*10000})
tbpc_bulk.epidish <- apply(tbpc_bulk.epidish, 2, function(x){(x/sum(x))*10000})
tbpc_epidish <- epidish(beta.m = (as.matrix(tbpc_bulk.epidish)),
                       ref.m = as.matrix(Derisoud.epidish),
                       method = "RPC")
summary(tbpc_epidish$estF)
```

```{r}
round(apply(tbpc_epidish$estF, 2, mean),3)
round(apply(tbpc_epidish$estF, 2, sd),3)
```



The deconvolution results of tbpc microarray data using two different deconvolution methods
shows roughly consistent patterns.
Some CTB, EVT, More Stromal, Some perivascular.


However, a good thing about these deconvolution results is that there is not much
cell type abundance difference across samples. It means that adjusting for cell type
abundance in DE analyses is not really important.
























